# Case Analysis of SP500 

We will investigate data related to the S&P500 stocks.

This is the URL of the data we will use:
https://docs.google.com/spreadsheets/d/11ahk9uWxBkDqrhNm7qYmiTwrlSC53N1zvXYfv7ttOCM/edit#gid=0 


### 1) Reading S&P500 data from a Google Sheet into a tibble

The Google Sheet ID is: `11ahk9uWxBkDqrhNm7qYmiTwrlSC53N1zvXYfv7ttOCM`. 
We can use the function `gsheet2tbl` in package `gsheet` to read the Google Sheet into a dataframe, as demonstrated in the following code.

```{r, warning=FALSE}
# Read S&P500 stock data present in a Google Sheet.
library(gsheet)

prefix <- "https://docs.google.com/spreadsheets/d/"
sheetID <- "11ahk9uWxBkDqrhNm7qYmiTwrlSC53N1zvXYfv7ttOCM"
# Form the URL to connect to
url500 <- paste(prefix, sheetID) 
# Read the Google Sheet located at the URL into a tibble called sp500
sp500 <- gsheet2tbl(url500)
```


### 1) Reviewing the data
We run the str() function to better understand the data.
```{r, warning=FALSE}
str(sp500)
```

The `str(sp500)` output provides valuable insights into the structure and data types of the columns in the `sp500` tibble. Let's delve into the details:

The output reveals that `sp500` is a tibble with dimensions [503 × 36]. This means it consists of 503 rows, each representing a specific S&P500 stock, and 36 columns containing information about each stock.

Here's a breakdown of the information associated with each column:

- The columns labeled `Date`, `Stock`, `Description`, `Sector`, and `Industry` are character columns. They respectively represent the date, stock ticker symbol, description, sector, and industry of each S&P500 stock.

- Columns such as `Market.Capitalization`, `Price`, `X52.Week.Low`, `X52.Week.High`, and other numeric columns contain diverse financial metrics and stock prices related to the S&P500 stocks.

- The column labeled `Technical.Rating` is a character column that assigns a technical rating to each stock.

By examining the `str(sp500)` output, you gain a comprehensive understanding of the data types and column names present in the `sp500` tibble, enabling you to grasp the structure of the dataset effectively.

