# Continuous Data (2 of 3)

## Overview of Bivariate Continuous Data

1.  Reading Data and Attaching Data to Memory

```{r echo=TRUE, eval=TRUE}
data(mtcars)
attach(mtcars)
```

## Bivariate Continuous and Categorical data

1.  Bivariate Relationship between Weight (wt) and Transmission (am)

2.  Display a summary table showing the descriptive statistics of weight of the cars broken down by transmission (am=1 or am=0)

### aggregate()

```{r}
aggregate(mtcars$wt, 
          by = list("am" = mtcars$am), 
          mean)
aggregate(mtcars$wt, 
          by = list("am" = mtcars$am), 
          sd)
```

### tapply()

```{r}
tapply(mtcars$wt, mtcars$am, mean)
tapply(mtcars$wt, mtcars$am, sd)
```

## Visualizing Means -- mean plot showing the average weight of the cars, broken down by transmission (am=1 & am=0)

```{r}
library(gplots)
plotmeans(wt ~ am
          ,data = mtcars
          ,mean.labels = TRUE
          ,digits=3
          ,main = "Mean (wt) by am = {0,1} "
          )
```

## Visualizing Median using Box Plot -- median weight of the cars broken down by transmission (am=1 & am=0)

```{r}
boxplot(wt~am
        , xlab = "am"
        , ylab = "Weight"
        , horizontal = TRUE
        )
```

## Bivariate Relationship between Weight (wt) and Cylinders (cyl)

Display a summary table showing the mean weight of the cars broken down by cylinders (cyl=4,6,8)

```{r}
psych::describeBy(wt, cyl)
```

## Show a mean plot showing the mean weight of the cars broken down by cylinders (cyl=4,6,8)

```{r}
library(gplots)
plotmeans(wt ~ cyl, 
          data = mtcars
          , mean.labels = TRUE
          , digits=2
          , main = "Mean (wt) by cyl = {4,6,8} ")
```

## Show a box plot showing the median weight of the cars broken down by cylinders (cyl=4,6,8)

```{r}
boxplot(wt ~ cyl, 
        xlab = "cyl", ylab = "Weight"
        )
```

## Distribution of Weight (wt) by Cylinders (cyl = {4,6,8}) and Transmisson Type (am = {0,1})

```{r}
aggregate(wt, 
          by = list("am" =am, "cyl" = cyl),
          mean)
```

## Visualization - Show a box plot showing the mean weight of the cars broken down by Transmission Type (am=1 & am=0) & cylinders (cyl=4,6,8)

```{r}
boxplot(wt ~ am:cyl
        , xlab = "cyl"
        , ylab = "Weight"
        )
```

## Visualization - Show a mean plot showing the mean weight of the cars broken down by Transmission Type (am=1 & am=0) & cylinders (cyl=4,6,8)

```{r warning=FALSE, message=FALSE}
library(gplots)
plotmeans(wt ~ interaction(am, cyl, sep = ", ")
          , data = mtcars
          , mean.labels = TRUE
          , digits=2
          , connect = FALSE
          , main = "Mean (wt) by cyl = {4,6,8} & am = {0,1}"
          , xlab= "cyl = {4,6,8} & am = {0,1}"
          , ylab="Average Weight"
          )
```
