# Continuous Data (2 of 4)

*Aug 1, 2023. V3.1 -=- This chapter is being heavily edited; It is very much Work in Progress*

## Using ggplot2 for Visualizing Univariate Continuous Data

1. In R, we can create these plots using the `ggplot2` package. For instance, the function `geom_boxplot()` produces box plots, `geom_violin()` creates violin plots, and `geom_histogram()` and `geom_density()` generate histograms and density plots, respectively. The `ggbeeswarm` package can be used for creating bee swarm plots with the `geom_beeswarm()` function. We will go into more detail about how to use these functions and interpret these plots in the upcoming sections. 



2. **Data**: Let us work with the same mtcars data from the previous chapter. Suppose we have run the following code:

```{r, warning=FALSE}
# Load the required libraries, suppressing annoying startup messages
library(tibble) 
suppressPackageStartupMessages(library(dplyr))
# Read the mtcars dataset into a tibble called tb
data(mtcars)
tb <- as_tibble(mtcars)
attach(tb)
# Convert several numeric columns into factor variables
tb$cyl <- as.factor(tb$cyl)
tb$vs <- as.factor(tb$vs)
tb$am <- as.factor(tb$am)
tb$gear <- as.factor(tb$gear)
```


### Rug Plot using ggplot

Rug plots are a one-dimensional representation of data where a small vertical line (or a 'tick') is drawn for each data point along the x-axis. They are typically used alongside other plots to better show the distribution of data points (Hassell, 2001). In R, the geom_rug() function from the ggplot2 package creates rug plots.

```{r}
library(ggplot2)
ggplot(tb, aes(x=wt)) + geom_rug()
```


## References

[1]
Hassell, N. (2001). Rug plots. Significance, 44(4), 181-182.
