# Continuous Data (5 of 6)

*Aug 4, 2023.*

## Exploring bivariate Continuous x Continuous data

THIS CHAPTER explores how to summarize and visualize the interaction between *bivariate continuous data* using correlation analysis, scatter plots, scatter plot matrices and other such techniques.

**Data**: Let us work with the same `mtcars` data from the previous chapter. Suppose we run the following code to prepare the data for subsequent analysis. The data is now in a tibble called `tb`:

```{r, warning=FALSE}
# Load the required libraries, suppressing annoying startup messages
library(tibble) 
suppressPackageStartupMessages(library(dplyr))
# Read the mtcars dataset into a tibble called tb
data(mtcars)
tb <- as_tibble(mtcars)
# Convert several numeric columns into factor variables
tb$cyl <- as.factor(tb$cyl)
tb$vs <- as.factor(tb$vs)
tb$am <- as.factor(tb$am)
tb$gear <- as.factor(tb$gear)
# Directly access the data columns of tb, without tb$mpg
attach(tb)
```

## Scatterplots

A scatter plot is a type of graph used to display the relationship between two continuous variables. It is a graphical representation of a bivariate distribution, where the values of two variables are plotted as points on a two-dimensional coordinate system.

A scatter plot can be used to identify trends, clusters, outliers, and other patterns in the data. It is also useful for detecting the presence of any outliers or influential observations that may affect the analysis.

The mtcars data set in R is a built-in data set that contains data on various car models. To create a scatter plot of mpg (miles per gallon) against wt (weight) in the mtcars data set, you can use the following code:

### Scatterplot using plot()

```{r}
plot(tb$wt, 
     tb$mpg, 
     main = "Scatter Plot of MPG vs. Weight",
     xlab = "Weight", ylab = "MPG", pch = 16)

```

This code will first load the mtcars data set, then create a scatter plot of `mpg` against `wt` using the `plot()` function. The main argument adds a title to the plot, the `xlab` and `ylab` arguments add axis labels, and the pch argument changes the shape of the points to a solid circle. The resulting scatter plot will show the relationship between `mpg` and `wt` in the `mtcars` data set.

## Scatterplot Matrix

A scatter plot matrix (also called a pairs plot or a SPLOM) is a graphical display of pairwise scatter plots of a set of variables. In a scatter plot matrix, each variable in the dataset is plotted against every other variable in a matrix format. This allows us to visualize the relationships between pairs of variables and explore potential patterns or trends in the data.

A scatter plot matrix is particularly useful for exploring multivariate datasets, as it allows us to quickly identify which pairs of variables may be strongly correlated, which may have weak or no correlation, and which may exhibit nonlinear relationships. It can also be used to identify outliers or unusual observations, and to visualize clusters or groups of observations based on patterns in the scatter plots.

### Scatterplot Matrix Using pairs()

```{r}

# scatter plot matrix for mpg, wt, hp, drat
pairs(tb[,c("mpg","wt","hp","drat")], pch = 19)

```

### Scatterplot Matrix Using scatterplotMatrix()

```{r warning=FALSE,message=FALSE}
# Load the car package
library(car)

# Create a scatterplot matrix using scatterplotMatrix()
scatterplotMatrix(~ mpg + disp +drat +wt, 
                  data = tb, col = c("blue", "red")) 

```

### Scatterplot Matrix Using pairs.panels()

```{r warning=FALSE,message=FALSE}
# Load the psych package
library(psych)

# Create a scatterplot matrix using pairs.panels()
pairs.panels(tb[,c("mpg","wt","hp","drat")], 
             main = "Scatterplot Matrix")

```
